Concentration of Maxima and Fundamental Limits in High-Dimensional Testing and Inference

### 2021-01-13 2.background/2.statistical_risks.tex
edit up to Chapter 2; leaving some for discussion.

We define the statistical risk metrics for signal detection and signal support recovery problems in this section.

{\bf Signal detection.}
Recall that in sparse signal detection problems, our goal is to come up with a procedure, $\mathcal{R}(x)$, such that the null hypothesis is rejected if the data $x$ is deemed incompatible with the null.
In the example of additive error models \eqref{eq:model-additive}, we wish to tell apart two hypotheses
\begin{equation} \label{eq:global-test-additive}
    \mathcal{H}_0: \mu(i) = 0, \;i=1,\ldots,p,
    \quad\text{v.s.}\quad 
    \mathcal{H}_1: \mu(i)\neq 0, \; \text{for some }i\in\{1,\ldots,p\},
\end{equation}
based on the $p$-dimensional observation $x$.
Similarly in the chi-square model \eqref{eq:model-chisq}, we look to test
\begin{equation} \label{eq:global-test-chisq}
    \mathcal{H}_0: \lambda(i) = 0, \;i=1,\ldots,p,
    \quad\text{v.s.}\quad 
    \mathcal{H}_1: \lambda(i)\neq 0, \; \text{for some }i\in\{1,\ldots,p\}.
\end{equation} 
Since the decision is binary, we may write the outcome of the procedure in the form of an indicator function, $T(\mathcal{R}(x))\in\{0,1\}$, where the function $T$ takes on value 1 if the null is to be rejected in favor of the alternative, and 0 if we fail to reject the null. 
The Type I and Type II errors of the procedure, i.e., the probability of wrong decisions under the null hypothesis $\mathcal{H}_0$ and alternative hypothesis $\mathcal{H}_1$, respectively, are defined as 
\begin{equation} \label{eq:type-II-error}
    \alpha(\mathcal{R}) := \P_{\mathcal{H}_0}\left(T(\mathcal{R}(x))=1\right)
    \quad \text{and} \quad
    \beta(\mathcal{R}) := \P_{\mathcal{H}_1}\left(T(\mathcal{R}(x))=0\right).
\end{equation}
The Neyman-Pearson framework of hypothesis testing then seeks tests that minimize the Type II error of the test, while controlling the Type I error of the test at low levels.
We are particularly interested in the sum of the two errors, 
\begin{equation} \label{eq:risk-detection}
    \mathrm{risk}^{\mathrm{D}}(\mathcal{R}) := \alpha(\mathcal{R}) + \beta(\mathcal{R}),
\end{equation}
which shall be referred to as the risk of signal detection (of the procedure $\mathcal{R}$).
It is trivial that a small $\mathrm{risk}^{\mathrm{D}}$ would imply both small Type I and Type II errors of the procedure.

@@

{\bf Signal support recovery.}
Turning to support recovery problems, our goal is to design a procedure $\mathcal R$ that produces a set estimate $\widehat{S}$ of the true index set of relevant variables $S$.
% The true index set, of course, depends on the modeling assumptions. 
For example, in the sparse additive error model \eqref{eq:model-additive} we aim to estimate $S=\{i:\mu(i)\neq 0\}$, while in the sparse chi-square model \eqref{eq:model-chisq} the goal is to estimate $S=\{i:\lambda(i)\neq 0\}$.
Formally, one should write $\widehat{S}(\mathcal{R}(x))$ to reflect the dependence of the set estimate on the procedure $\mathcal{R}$ and on the test statistics $x$; 
for notational convenience, we suppress this dependence and simply write $\widehat{S}$.
% in place of $\widehat{S}(\mathcal{R}(x))$

For a given procedure $\mathcal{R}$, the {false discovery rate} (FDR) 
and the {false non-discovery rate} (FNR) of a procedure are defined as
% of the procedure is defined to be the expected fraction of false findings not in the true index set, among the reported discoveries \cite{benjamini1995controlling}. 
% Its counterpart, \emph{false non-discovery rate} (FNR), measuring the power of the procedure, is defined as the expected fraction of missed detection. 
% Mathematically, we define
\begin{equation} \label{eq:FDR-FNR}
    \mathrm{FDR}(\mathcal{R}) := \E\left[\frac{|\widehat{S}\setminus S|}{\max\{|\widehat{S}|,1\}}\right]
    \quad \text{and} \quad
    \mathrm{FNR}(\mathcal{R}) := \E\left[\frac{|S\setminus \widehat{S}|}{\max\{|{S}|,1\}}\right],
\end{equation}
where the maxima in the denominators resolve the possible division-by-0 problem. 
Roughly speaking, FDR measures the expected fraction of false findings, while FNR describes the proportion of Type II errors among the true signals, and reflects the average marginal power of the procedure.

A more stringent criterion for false discovery is the \ac{FWER},
%family-wise error rate (FWER),
% defined to be the probability of reporting at least one finding not contained in the true index set.
and correspondingly, a more stringent criteria for false non-discovery is the \ac{FWNR},
%family-wise non-discovery rate (FWNR), 
i.e.,
% the probability of missing at least one signal in the true index set. That is,
\begin{equation} \label{eq:FWER-FWNR}
    \mathrm{FWER}(\mathcal{R}) := 1 - \P[\widehat{S} \subseteq S]
    \quad \text{and} \quad
    \mathrm{FWNR}(\mathcal{R}) := 1 - \P[S \subseteq \widehat{S}].
\end{equation}

We introduce five different statistical risk metrics, 
each having different asymptotic limits in the support recovery problems as we will see later in this text. 
Following \cite{arias2017distribution}, we define the risk for \emph{approximate} support recovery as
\begin{equation} \label{eq:risk-approximate}
    \mathrm{risk}^{\mathrm{A}}(\mathcal{R}) := \mathrm{FDR}(\mathcal{R}) + \mathrm{FNR}(\mathcal{R}).
\end{equation}
Analogously, we define the risk for \emph{exact} support recovery as
\begin{equation} \label{eq:risk-exact}
    \mathrm{risk}^{\mathrm{E}}(\mathcal{R}) := \mathrm{FWER}(\mathcal{R}) + \mathrm{FWNR}(\mathcal{R}).
\end{equation}
Two closely related measures of success in the exact support recovery risk are the probability of exact recovery, 
\begin{equation} \label{eq:risk-prob}
    \P[\widehat{S} = S] = 1 - \P[\widehat{S} \neq S],
\end{equation}
and the Hamming loss
\begin{equation} \label{eq:Hamming-loss}
    H(\widehat S, S) := \left|\widehat{S}\triangle S\right|
    = \sum\limits_{i=1}^p\left|\mathbbm{1}_{\widehat{S}}(i)- \mathbbm{1}_{S}(i)\right|.
\end{equation}
which counts the number of mismatches between the estimated and true support sets.

@@

Notice that all risk metrics introduced so far penalize false discoveries and missed signals somewhat symmetrically --- the approximate support recovery risk combines proportions of errors, the exact support recovery risk combines probabilities of errors, and the Hamming loss increments the risk by one regardless of the types of errors made.
In applications, however, attitudes towards Type I and Type II errors are often different.
In the example of GWAS, where the number of candidate locations $p$ could be in the millions, researchers are typically interested in the marginal (location-wise) power of discovery, while exercising stringent (family-wise) false discovery control --- a situation not reflected in the above-mentioned risk metrics.
This asymmetric consideration, and in particular the GWAS application, prompts us to consider risks that weigh both the family-wise error rate and the marginal power of discovery.
One such risk metric is what shall be referred to as the \emph{exact-approximate} support recovery risk
\begin{equation} \label{eq:risk-exact-approx}
    \mathrm{risk}^{\mathrm{EA}}(\mathcal{R}) := \mathrm{FWER}(\mathcal{R}) + \mathrm{FNR}(\mathcal{R}).
\end{equation}
The somewhat cumbersome name and notation are chosen to reflect
the asymmetry in dealing with the two types of errors in support recovery.
Namely, when the risk metric \eqref{eq:risk-exact-approx} vanishes, we have ``exact false discovery control, and approximate false non-discovery control'' asymptotically.

>>>>

  
We define the statistical risk metrics for signal detection and signal support recovery problems in this section.
Formally, we denote a statistical procedure, i.e., measurable function of the data, as ${\cal R}={\cal R}(x)$.
In the testing context, a procedure ${\cal R}$ produces a binary decision $T$ that represents our judgment on the presence or absence of a signal. 
In the support recovery problem, a procedure ${\cal R}$ produces an index set $\widehat{S}$ that represents our estimate of the signal support.  
The statistical risks are then suitable functionals of ${T}$ and $\widehat{S}$ in respective contexts. 
%As we will see, these risk metrics will range from the familiar Type I, Type II errors in detection problems, to various mis-classification errors.

{\bf Signal detection.} 
Recall that in sparse signal detection problems, our goal is to come up with a procedure, $\mathcal{R}(x)$, such that the null hypothesis is rejected if the data $x$ is deemed incompatible with the null.
In the additive error models context \eqref{eq:model-additive}, we wish to tell apart two hypotheses
\begin{equation} \label{eq:global-test-additive}
    \mathcal{H}_0: \mu(i) = 0, \;i=1,\ldots,p,
    \quad\text{v.s.}\quad 
    \mathcal{H}_1: \mu(i)\neq 0, \; \text{for some }i\in\{1,\ldots,p\},
\end{equation}
based on the $p$-dimensional observation $x$.
Similarly, in the chi-square model \eqref{eq:model-chisq}, we look to test
\begin{equation} \label{eq:global-test-chisq}
    \mathcal{H}_0: \lambda(i) = 0, \;i=1,\ldots,p,
    \quad\text{v.s.}\quad 
    \mathcal{H}_1: \lambda(i)\neq 0, \; \text{for some }i\in\{1,\ldots,p\}.
\end{equation} 
Since the decision is binary, we may write the outcome of the procedure in the form of an indicator function, 
$T(\mathcal{R}(x))\in\{0,1\}$, where $T=1$ if the null is to be rejected in favor of the alternative, and 0 if we fail to reject the null. 
The Type I and Type II errors of the procedure, i.e., the probability of wrong decisions under the null hypothesis $\mathcal{H}_0$ and alternative hypothesis $\mathcal{H}_1$, respectively, are defined as 
\begin{equation} \label{eq:type-II-error}
    \alpha(\mathcal{R}) := \P_{\mathcal{H}_0}\left(T()\mathcal{R}(x))=1\right)
    \quad \text{and} \quad
    \beta(\mathcal{R}) := \P_{\mathcal{H}_1}\left(T(\mathcal{R}(x))=0\right).
\end{equation}
The Neyman-Pearson framework of hypothesis testing then seeks tests that minimize the Type II error of the test, while controlling the Type I error of the test at low levels.
We are particularly interested in the sum of the two errors, 
\begin{equation} \label{eq:risk-detection}
    \mathrm{risk}^{\mathrm{D}}(\mathcal{R}) := \alpha(\mathcal{R}) + \beta(\mathcal{R}),
\end{equation}
which shall be referred to as the risk of signal detection (of the procedure $\mathcal{R}$).
It is trivial that a small $\mathrm{risk}^{\mathrm{D}}$ would imply both small Type I and Type II errors of the procedure.

@@

{\bf Signal support recovery.}
Turning to support recovery problems, our goal is to design a procedure that produces a set estimate $\widehat{S}({\cal R}(x))$ of the true index set of relevant variables $S$.
% The true index set, of course, depends on the modeling assumptions. 
For example, in the sparse additive error model \eqref{eq:model-additive} we aim to estimate $S=\{i:\mu(i)\neq 0\}$, while in the sparse chi-square model \eqref{eq:model-chisq} the goal is to estimate $S=\{i:\lambda(i)\neq 0\}$.
For simplicity of notation, we shall write $\widehat{S}$ for the support estimator $\widehat{S}({\cal R}(x))$.

For a given procedure $\mathcal{R}$, its {false discovery rate} (FDR) 
and  {false non-discovery rate} (FNR) are defined, respectively, as
% of the procedure is defined to be the expected fraction of false findings not in the true index set, among the reported discoveries \cite{benjamini1995controlling}. 
% Its counterpart, \emph{false non-discovery rate} (FNR), measuring the power of the procedure, is defined as the expected fraction of missed detection. 
% Mathematically, we define
\begin{equation} \label{eq:FDR-FNR}
    \mathrm{FDR}(\mathcal{R}) := \E\left[\frac{|\widehat{S}\setminus S|}{\max\{|\widehat{S}|,1\}}\right]
    \quad \text{and} \quad
    \mathrm{FNR}(\mathcal{R}) := \E\left[\frac{|S\setminus \widehat{S}|}{\max\{|{S}|,1\}}\right],
\end{equation}
where the maxima in the denominators resolve the possible division-by-0 problem. 
Roughly speaking, FDR measures the expected fraction of false findings, while FNR describes the proportion of Type II errors among the true signals, and reflects the average marginal power of the procedure.

A more stringent criterion for false discovery is the family-wise error rate (\ac{FWER}), defined to be the probability of 
reporting at least one finding not contained 
in the true index set. Correspondingly, a more stringent criterion for false non-discovery is the family-wise non-discovery rate (\ac{FWNR}), i.e., the probability of missing at least one signal in the true index set. That is,
\begin{equation} \label{eq:FWER-FWNR}
    \mathrm{FWER}(\mathcal{R}) := 1 - \P[\widehat{S} \subseteq S]
    \quad \text{and} \quad
    \mathrm{FWNR}(\mathcal{R}) := 1 - \P[S \subseteq \widehat{S}].
\end{equation}

We introduce five different statistical risk metrics, 
each having different asymptotic limits in the support recovery problems as we will see in 
Chapter \ref{chap:phase-transitions}.
Following \cite{arias2017distribution}, we define the risk for \emph{approximate} support recovery as
\begin{equation} \label{eq:risk-approximate}
    \mathrm{risk}^{\mathrm{A}}(\mathcal{R}) := \mathrm{FDR}(\mathcal{R}) + \mathrm{FNR}(\mathcal{R}).
\end{equation}
Analogously, we define the risk for \emph{exact} support recovery as
\begin{equation} \label{eq:risk-exact}
    \mathrm{risk}^{\mathrm{E}}(\mathcal{R}) := \mathrm{FWER}(\mathcal{R}) + \mathrm{FWNR}(\mathcal{R}).
\end{equation}
Two closely related measures of success in the exact support recovery risk are the probability of exact recovery, 
\begin{equation} \label{eq:risk-prob}
    \P[\widehat{S} = S] = 1 - \P[\widehat{S} \neq S],
\end{equation}
and the Hamming loss
\begin{equation} \label{eq:Hamming-loss}
    H(\widehat S, S) := \left|\widehat{S}\triangle S\right|
    = \sum\limits_{i=1}^p\left|\mathbbm{1}_{\widehat{S}}(i)- \mathbbm{1}_{S}(i)\right|.
\end{equation}
which counts the number of mismatches between the estimated and true support sets.

@@

Notice that all risk metrics introduced so far penalize false discoveries and missed signals somewhat symmetrically --- the approximate support recovery risk combines proportions of errors, the exact support recovery risk combines probabilities of errors, and the Hamming loss increments the risk by one regardless of the types of errors made.
In applications, however, attitudes towards Type I and Type II errors are often different.
In the example of GWAS, where the number of candidate locations $p$ could be in the millions, researchers are typically interested in the marginal (location-wise) power of discovery, while exercising stringent (family-wise) false discovery control. 
Such types of asymmetric risks, while important in applications, have not been studied theoretically.  
This asymmetric consideration, and in particular the GWAS application, prompts us to consider risks that weigh both the family-wise error rate and the marginal power of discovery.
One such risk metric is what shall be referred to as the \emph{exact-approximate} support recovery risk
\begin{equation} \label{eq:risk-exact-approx}
    \mathrm{risk}^{\mathrm{EA}}(\mathcal{R}) := \mathrm{FWER}(\mathcal{R}) + \mathrm{FNR}(\mathcal{R}).
\end{equation}
The somewhat cumbersome name and notation are chosen to reflect
the asymmetry in dealing with the two types of errors in support recovery.
Namely, when the risk metric \eqref{eq:risk-exact-approx} vanishes, we have ``exact false discovery control, and approximate false non-discovery control'' asymptotically.

### 2021-01-13 6.GWAS/6.proofs-GWAS.tex
edit up to Chapter 2; leaving some for discussion.

The proof follows the ideas in \cite{arias2017distribution}, and is very similar to the proof of Theorem \ref{thm:Gaussian-error-approx-boundary}. 
One could in principle obtain the proofs in this section by referencing arguments that have appeared in Section \ref{sec:proofs}.
We choose to present the the proof here in full to make this section self-contained.

@@

As with the proof of Theorem \ref{thm:chi-squared-approx-boundary}, one could shorten the presentations in this section by referencing arguments Section \ref{sec:proofs}.
Again, we choose to present the the proof in full to make this section self-contained.

>>>>

The proof follows the ideas in \cite{arias2017distribution}, and is very similar to the proof of Theorem \ref{thm:Gaussian-error-approx-boundary}. 
One could in principle obtain the proofs in this section by referencing arguments that have appeared in Section \ref{sec:proofs}.
We choose to present the proof here in full to make this section self-contained.

@@

As with the proof of Theorem \ref{thm:chi-squared-approx-boundary}, one could shorten the presentations in this section by referencing arguments Section \ref{sec:proofs}.
Again, we choose to present the proof in full to make this section self-contained.
